---
title: "W203 Lab 2 Final: Introduction and model building process"
output:
  pdf_document: default
  html_notebook: default
---

```{r, warning=FALSE, echo=FALSE}
#Install packages and load libraries 

#Pastecs for nice, quick summary features of our datasets. 
#install.packages("pastecs", repos = "http://cran.us.r-project.org")

library("readxl")
library("tidyverse")
library("openxlsx")
library(dplyr)
library(zoo)
library(grid)
library(gridExtra)
library(corrplot)
```

```{r}
mobility_report <- read.csv("~/w203_lab2_group1/data/raw/mobility_report.csv")
covid_cases <- read.csv("~/w203_lab2_group1/data/raw/covid_cases.csv")
df <- read.csv("~/w203_lab2_group1/data/cleaned/df.csv")
```

## Introduction

It has been over a year since the initial outbreak of COVID-19 in the US. During this year, many states have adopted various policies to slow down the spread of the pandemic. For example, in New York city, Gov. Andrew Cuomo shut indoor on March 16, 2020. Many companies also started to operate in telecommuting mode. As a result, people’s mobility pattern is very different from the pre COVID-19 period. This presents an opportunity for studying the impact of these mobility changes on the spreading of COVID-19.

We think this is an interesting research question now because some states are starting to gradually revert some policies. For example, as of 3/19, indoor dining can increase to 50% capacity in New York City. The research result of the effects of change of mobility on new covid cases would help people and states make decisions on whether to increase their retail and recreation activity.
We propose exploring a causal relationship between the mobility, public mask mandate and demographics within a state and COVID-19 cases. Formally, our primary research question is: 

How much does a percentage change in mobility impact the percentage change in new COVID-19 cases within a given state, given a 2 week lag period? 

We choose to use 2 weeks between mobility & mask mandate and new COVID-19 cases based on the current CDC guidelines [@CDC] suggesting an individual is likely to incubate the disease 2-14 days after initial symptoms appear (which could be up to 10 days after initial infection).Thus 2 weeks after (i.e. days 14 to 21) mobility/mask mandate best covers this period of 12 to 24 days. 

## Model Building Process

### Data Sources

* New York Times COVID-19 Cases Data [@NYT]: A series of data files with cumulative counts of coronavirus cases in the United States, at the state and county level, over time.
* COVID-19 US State Policy Database [@CUSP]: A database of state policy responses to the pandemic, compiled by researchers at the Boston University School of Public Health.
* COVID-19 Community Mobility Report [@Google]: A Google dataset that includes state-level measurements of individual mobility
* The American Community Survey [@USCensus]: A product from the US Census Bureau that contains state-level demographics and other indicators of general interest.

### Choosing the variables

In this section we will define the dependent variable and independent variables, as well as our reasons to choose them. 

#### Choosing the time period

We decide to use mobility variables during the week June 28th, 2020 to July 4th, 2020 and the new covid cases variable with a 2 week lag - July 18th, 2020 to Jult

#### Dependent variable

##### percent_change_in_week_over_week_covid_cases (July 18th - 24th)

Percentage change in new COVID cases, comparing the number of new cases from Sunday 07/19/20 - Saturday 07/25/20 vs. Sunday 07/12/20 - Saturday 07/18/20.

The time period of the dependent variable is 2 weeks lagged from the mobility variable. The time period of the dependent variable is 2 weeks lagged from the mobility variable. We choose the 2 weeks lagging period based on the CDC suggestion that an individual is likely to incubate the disease 2-14 days after initial symptoms appear (which could be up to 10 days after initial infection).

We chose to use this time period for the following reasons

* This time period captures the return towards baseline mobility after the initial lock down dip (quarantine fatigue) which is an interesting time period to study (Figure 2) 
* This is a week in which 31 states did not have public face mask mandates vs 20 states did, and therefore is interesting time period to study the effects of public face mask mandates on slowing down the spread of COVID-19.
* Is a counter to the seasonal nature of the disease, i.e. weather plays less of a role in transmission during this summer time period which helps our causal question.
* It's early enough in the pandemic that we'd expect a much lower proportion of the population to have contracted the disease already and develop antibodies, which gives us a clearer picture of our causal question.

Our method of choosing percent_change and the lag between change in covid cases and change in mobility is akin to the “differenced panels” approach on async video 11.14. Similar to “differenced panels approach”. By taking the difference of percent in new covid cases, we can reduce some impacts of the confounding variables.

#### Independent variables

##### percent_change_in_week_over_week_mobility_retail_and_recreation (Sat,Sun,M...F)

Seven variables, each a percentage point change in median mobility (compared to baseline) for Retail & Recreation for the week of Sunday 06/28/20 - Saturday 07/04/20  vs Sunday 06/21/20 - Saturday 06/27/20 for each day of the week. Sunday is only compared to Sunday, etc.

Since we have no indication of how to weight these values based on what the real baseline number is, we can’t simply average all 7 days. For example, a -10 percent change of mobility on Monday is very different from a -10 percent change of mobility on Saturday since the baseline number of people for Saturday is much higher. Therefore we decide to keep these variables separate. 

##### public_mask_mandate_flag (In place as of June 28th)

Flag for whether or not a public face mask mandate was in place for the state as of June 28th (i.e. the majority of the week of July 4th, our primary week of interest for change in mobility)


##### percent_female (2019)

American Community Survey (2019) estimate of state’s percentage of female residents.

This study [@HBS] suggests that "women are much more likely than men to view COVID-19 as a severe health problem. They are also more willing to wear face masks and follow other public health recommendations to prevent the spread of the virus countries."

Therefore we decide to include this variable as a control variable.

##### percent_white (2019)

American Community Survey (2019) estimate of state’s percentage of white residents.

This article [@MayoClinic] suggests that "Research increasingly shows that racial and ethnic minorities are disproportionately affected by coronavirus disease 2019 (COVID-19) in the United States."

Therefore we decide to include this variable as a control variable.

##### percent_65_and_older (2019)

American Community Survey (2019) estimate of state’s percentage of residents 65 or older (66 or older at time of analysis).

CDC suggests that "Older adults are At greater risk of requiring hospitalization or dying if diagnosed with COVID-19" [@CDC2].

Therefore we think that states with a high percentage of older adult population could have lower mobility due to the fear of getting COVID-19 and therefore, lead to lower COVID-19 cases.We decide to include this variable as a control variable.

##### percent_24_and_younger (2019)

American Community Survey (2019) estimate of state’s percentage of residents 24 or younger (25 or younger at time of analysis) .

According to the news, some college students have shown reckless behavior with respect to social distancing [@InsideHigherEd]. 

Therefore we decide to include this variable as a control variable.

### Choosing the mobility variable

```{r, warning=FALSE, echo=FALSE}
get_plots <- function(mobility, covid_cases, state_var) {

mobility_report <- mobility

covid_cases$date <- as.Date(covid_cases$date)
# Add new cases
covid_cases <- covid_cases %>% group_by(fips) %>% mutate(new_cases=c(NA, diff(cases)))
  
mob_ny <- mobility_report %>% mutate(date=as.Date(date)) %>% filter(sub_region_1==state_var)  %>% filter(sub_region_2=='')
mob_ny_smooth <- mob_ny %>% mutate(
  retail_and_recreation=rollmean(x=retail_and_recreation_percent_change_from_baseline, 7, na.pad=TRUE, align="right"),
  grocery_and_rx=rollmean(x=grocery_and_pharmacy_percent_change_from_baseline, 7, na.pad=TRUE, align="right"),
  transit=rollmean(x=transit_stations_percent_change_from_baseline, 7, na.pad=TRUE, align="right"),
  park=rollmean(x=parks_percent_change_from_baseline, 7, na.pad=TRUE, align="right"),
  residence=rollmean(x=residential_percent_change_from_baseline, 7, na.pad=TRUE, align="right"),
  workplaces=rollmean(x=workplaces_percent_change_from_baseline, 7, na.pad=TRUE, align="right"),
)

covid_ny <- covid_cases %>% filter(state==state_var) %>% mutate(new_cases_smooth=rollmean(x=new_cases, 7, na.pad=TRUE, align="right"))
joined <- left_join(covid_ny, mob_ny_smooth, by='date')

retail_plot <- joined %>% ggplot(aes(x=date)) + geom_line(aes(y=retail_and_recreation)) + 
  geom_line(aes(y=new_cases_smooth / 1000), color='green') + scale_x_date(date_labels = "%m") + xlab("month")

grocery_plot <- joined %>% ggplot(aes(x=date)) + geom_line(aes(y=grocery_and_rx)) + 
  geom_line(aes(y=new_cases_smooth / 1000), color='green') + scale_x_date(date_labels = "%m") + xlab("month")

transit_plot <- joined %>% ggplot(aes(x=date)) + geom_line(aes(y=transit)) + 
  geom_line(aes(y=new_cases_smooth / 1000), color='green') + scale_x_date(date_labels = "%m") + xlab("month")

park_plot <- joined %>% ggplot(aes(x=date)) + geom_line(aes(y=park)) + 
  geom_line(aes(y=new_cases_smooth / 1000), color='green') + scale_x_date(date_labels = "%m") + xlab("month")

resid_plot <- joined %>% ggplot(aes(x=date)) + geom_line(aes(y=residence)) + 
  geom_line(aes(y=new_cases_smooth / 1000), color='green') + scale_x_date(date_labels = "%m") + xlab("month")

work_plot <- joined %>% ggplot(aes(x=date)) + geom_line(aes(y=workplaces)) + 
  geom_line(aes(y=new_cases_smooth / 1000), color='green') + scale_x_date(date_labels = "%m") + xlab("month")

plots <- grid.arrange(
  retail_plot,
  grocery_plot,
  transit_plot,
  park_plot,
  resid_plot,
  work_plot,
  nrow = 3,
  top = "Figure 2: Moblity and New Covid Cases in California\n",
  bottom = "Mobility(Black): percent\nNew Covid Cases(Green): 1000 cases"
)

return(plots)
}
```

In this section, we will show the process of choosing the mobility variable. The two charts show (1) the correlation plot between different variables and (2) the time series plot show the relationship between different mobility variables by time.

```{r, warning=FALSE, echo=FALSE}
M <- cor(mobility_report[, c(10: 15)] %>% rename(retail_and_recreation=retail_and_recreation_percent_change_from_baseline, grocery_and_rx=grocery_and_pharmacy_percent_change_from_baseline, parks=parks_percent_change_from_baseline, transit=transit_stations_percent_change_from_baseline, workspaces=workplaces_percent_change_from_baseline, residential= residential_percent_change_from_baseline) %>% drop_na())

corrplot(M, method='circle', title="Figure 1 Mobility variable correlations", mar=c(0,0,1,0))
```


```{r, warning=FALSE, echo=FALSE}
plots <- get_plots(mobility_report, covid_cases, 'California')
```

Figure 1, 2 show that retail, transit and grocery have positive correlation of > 0.70 (figure 2) and they tend to move in the same direction (figure 1). Retail and residential has a negative correlation of -0.75  and they tend to move approximately in the opposite direction.

Since retail mobility is positively correlated with transit and grocery and negatively correlated with residence mobility, we decide to only include one of them as our independent mobility variable. We decide to use retail mobility because we think its practical meaning is the most interesting among all the mobility variables, given the political focus on restaurants and storefronts. We decide not to use workplaces and park mobility variable because (1) park is outdoor and easier to maintain 6-feet social distances, and therefore, we don’t expect change in park mobility will cause change in new covid cases and (2) for workplaces, generally speaking people have less control of whether to go to workplaces or not, and therefore, this variable does not have sufficient practical impact.

Another observation is that there is no clear correlation between any of the mobility variables and new COVID-19 cases. Still, we decide to include mobility as an independent variable to study its combined effects with other independent variables (public mask mandate and demographic variables)

### Examine the control variables

In this section, we will examine the relationship between control variables (public_mask_mandate_flag, percent_female, percent_white, percent_65_and_older, percent_24_and_younger) and the dependent variable (percent_change_in_week_over_week_covid_cases) for the study time period (July 18th - 24th).

#### Public mask mandate

Looking the boxplot for percent change in COVID-19 cases group by public mask mandate variable for the 51 states (Figure 3), we see that there is a slight higher mean states without public mask mandate, but the variable is also pretty high to draw any conclusion from this plot. We decide to include this control variable and further examine its coefficient and p-value.

``` {r, warning=FALSE, echo=FALSE}
df %>% ggplot(aes(x=public_mask_mandate_flag, group=public_mask_mandate_flag, y=p1_p2_percent_change_in_cases)) + geom_boxplot() + ylab("% change in covid cases") + xlab("State-wide Public mask mandate (0 means no mandate)") + ggtitle("Figure 3: \n change in covid cases with/without public mask mandate")
```
#### Demongraphics variables

Here is a combined scatter plot for the four demographics variables we are interested in (percent_female, percent_white, percent_65_and_older, percent_24_and_younger). The scatter plots don't show a clear pattern of the linear relationship just by these control variables themselves. Nonetheless, we decide to include these variables in the model specification to study further.

```{r, warning=FALSE, echo=FALSE, message=FALSE}
percent_female <- df %>% ggplot(aes(x=percent_female, y=p1_p2_percent_change_in_cases)) + geom_point() + stat_smooth(se=TRUE) + xlab("% female population") + ylab("% change in covid cases")
percent_white <- df %>% ggplot(aes(x=percent_white, y=p1_p2_percent_change_in_cases)) + geom_point() + stat_smooth(se=TRUE) +xlab("% white population") + ylab("% change in covid cases")
percent_over_65 <- df %>% ggplot(aes(x=percent_over_65, y=p1_p2_percent_change_in_cases)) + geom_point() + stat_smooth(se=TRUE) +xlab("% population over 65 years old ") + ylab("% change in covid cases")
percent_under_24 <- df %>% ggplot(aes(x=percent_under_24, y=p1_p2_percent_change_in_cases)) + geom_point() + stat_smooth(se=TRUE) +xlab("% population under 24 years old ") + ylab("% change in covid cases")

grid.arrange(
  percent_female, percent_white, percent_over_65, percent_under_24, nrow = 2,
  bottom = "Figure 4: \nRelationship between demongraphics control variables and % change in covid cases\n"
)
```


### Modeling Goal

Our modeling goal is explanatory. Primarily, we are interested in the causal relationship between mobility change in retail and recreation of the week of 06/28/2020 and the change of new COVID-19 cases 2 weeks later. Secondarily, we are interested in the casual described above, with the control of public mask mandate and demographic variables as control variables.

We are only studying the casual relationship within the chosen time period and not planning to extend any conclusion beyond the original time period.

### Casual Theory

#### Increase in retail and recreation mobility causes increase in new covid cases

From CDC:
“The virus that causes COVID-19 most commonly spreads from person to person by respiratory droplets during close physical contact (within 6 feet). The virus can sometimes spread from person to person by small droplets or virus particles that linger in the air for minutes to hours.”

Based on this, our casual theory is that an increase in retail and recreation mobility would result in an increase in the number of people that each person is in contact with, and therefore, increase the chances of each individual infected by covid, and finally, result in an increase in new covid cases.

#### Public Facemask Mandate causes decrease in new covid cases

COVID-19 is airborne and spread by respiratory droplets which enter through the nose and mouth. Wearing a mask will help prevent the spread of these droplets into the air. It is possible for someone to spread the virus even if they do not have symptoms [@Olmsted].

Based on this, our casual theory is that public facemask mandate causes a decrease in new covid cases, so long as mobility is constant. In other words, public mask mandates moderate the impact mobility has on new COVID-19 cases. 

### Model Specifications

#### Linear Model

precent_change_in_week_over_week_covid_cases ~ precent_change_in_week_over_week_mobility_retail_and_recreation (Sunday, Monday, .. , Saturday)

#### Model Two

change_in_week_over_week_covid_cases ~ percent_change_in_week_over_week_mobility_retail_and_recreation (Sunday, Monday, .. , Saturday) + public_mask_mandate_flag

#### Model Three

percent_change_in_week_over_week_covid_cases ~ percent_change_in_week_over_week_mobility_retail_and_recreation (Sunday, Monday, .. , Saturday) + public_mask_mandate_flag + percent_white + percent_female + percent_under_24 + percent_over_65




